---
title: "Statistical Inference Course Project | Part 2"
author: "Onur Akpolat"
date: "22. November 2014"
output:
  html_document
---

#### 1. Load the ToothGrowth data and perform some basic exploratory data analyses.

```{r loading, echo=TRUE}
library(datasets)
x <- ToothGrowth
# convert dose to factor
x$dose <- as.factor(x$dose)
```

```{r exploratory, fig.height=3, fig.width=6, echo=TRUE}
str(x)
table(x$dose, x$supp)
library(ggplot2)
ggplot(data=x, aes(x=dose, y=len, fill=supp)) + geom_bar(stat="identity",) + facet_grid(. ~ supp) + xlab("Dose in miligrams") + ylab("Tooth length") + guides(fill=guide_legend(title="Supplement type"))
ggplot(aes(x=dose, y=len), data=x) + geom_boxplot(aes(fill=dose)) + xlab("Dose in miligrams") + ylab("Tooth length") + guides(fill=guide_legend(title="Dose"))
ggplot(aes(x=supp, y=len), data=x) + geom_boxplot(aes(fill=supp)) + xlab("Supplement type") + ylab("Tooth length") + guides(fill=guide_legend(title="Supplement type"))
```

#### 2. Provide a basic summary of the data.

```{r summary, echo=TRUE}
summary(x)
```

#### 3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering).

```{r confInterval, echo=TRUE}
t.test(len ~ supp, data = x)
```

Null hypothesis can not be rejected as confindence intervals contain zero and p-value is 0.06.

Supplement types seems to have no impact on Tooth growth.


```{r tests, echo=TRUE}
# three groups as per dose level pairs
x.doses_0.5_1.0 <- subset (x, dose %in% c(0.5, 1.0)) 
x.doses_0.5_2.0 <- subset (x, dose %in% c(0.5, 2.0)) 
x.doses_1.0_2.0 <- subset (x, dose %in% c(1.0, 2.0)) 

# Check for dose levels (0.5, 1.0)
t.test(len ~ dose, data = x.doses_0.5_1.0)

# Check for dose levels (0.5, 2.0)
t.test(len ~ dose, data = x.doses_0.5_2.0)

# Check for dose levels (1.0, 2.0)
t.test(len ~ dose, data = x.doses_1.0_2.0)
```

The p-value is less than 0.05 and confindence intervals don't contian 0. The average toot length increases with an inceasing dose. The null hypothesis can be rejected.

#### 4. State your conclusions and the assumptions needed for your conclusions.

Conclusions

1. Supplement type seem to have no impact on tooth growth.
2. Inreasing the dose level leads to increased tooth growth as well.

Assumptions

1. The experiment was done with random assignment of guinea pigs to different dose level categories and supplement type to control for confounders that might affect the outcome.

2. Members of the sample population, i.e. the 60 guinea pigs, are representative of the entire population of guinea pigs. This assumption allows us to generalize the results.

3. For the t-tests, the variances are assumed to be different for the two groups being compared. This assumption is less stronger than the case in which the variances are assumed to be equal.
